# Introduction to R


## About this module

This module will provide you with the fundamental skills in basic programming in R. 


- data wrangling

Upon the completion of this module, you will be able to load and preprocess data in R, write 

**Learning objectives**

- xxx
- xxx





## R programming language
    
R is a language that is applied in diverse fields of data science and analysis. Typical applications include...

* data wrangling
* statistical analysis
* machine learning
* data visualisation and maps
* processing spatial data
* geographic information analysis
* and many more.

Apart from its widespread use, there are a number of other reasons to learn R...

* R is free and open source.
* R has more comprehensive functionality than most proprietary solutions.
* R is avaialble for Windows, May and Linux
* R is a general-purpose programming language, so you can use it to automate analyses and create new custom functions that extend default features.
* Because R is open source, it has a large user community, so it is easy to get help.

R is a so called **high level programming language** or **scripting language**. This means that R code is not compiled into a computer readable format, but interpreted by an **interpreter.** An interpreter is a computer program that directly interprets and executes instructions written in a programming language.  

In order to make sure that the interpreter can understand the program code, the programmer must stick to the grammar of the programming language; i.e. the interpreter expects commands to appear in a predefined order. This grammar is often regarded as **Syntax**.  

In this lession we will focus on some key principles of the R syntax and logic. 

## Installation and Setup

Before you can run your code, you have to install R together with an **Integrated Development Environment IDE** on your machine: 

1. Download R from [R Archive Network (CRAN.)](http://cran.r-project.org){target="_blank"} 
2. Follow the instructions to and install the most up to date version on your machine (chose 'base' as well as 32-bit or 64-bit dependent on the bit-version of your operating system).

The IDE is where your write, test and execute your R programs. We strongly recommend using **RStudio Desktop**, which is [freely available for download.](https://www.rstudio.com/){target="_blank"}

```{block2, type = 'rmdtip'}
If you need help, please turn to the discussion forum!
```

## Schedule

The lectures and practical sessions have been designed to follow the schedule below

xxx


## Reference books

Suggested reading


- *Programming Skills for Data Science: Start Writing Code to Wrangle, Analyze, and Visualize Data with R* by Michael Freeman and Joel Ross, Addison-Wesley, 2019. See book [webpage](https://www.pearson.com/us/higher-education/program/Freeman-Programming-Skills-for-Data-Science-Start-Writing-Code-to-Wrangle-Analyze-and-Visualize-Data-with-R/PGM2047488.html) and [repository](https://programming-for-data-science.github.io/).
- *R for Data Science* by Garrett Grolemund and Hadley Wickham, O'Reilly Media, 2016. See [online book](https://r4ds.had.co.nz/).
- *Discovering Statistics Using R* by Andy Field, Jeremy Miles and Zoë Field, SAGE Publications Ltd, 2012. See book [webpage](https://www.discoveringstatistics.com/books/discovering-statistics-using-r/).
- *Machine Learning with R: Expert techniques for predictive modeling* by Brett Lantz, Packt Publishing, 2019. See book [webpage](https://subscription.packtpub.com/book/big_data_and_business_intelligence/9781788295864).

Further reading

- *The Art of R Programming: A Tour of Statistical Software Design* by Norman Matloff, No Starch Press, 2011. See book [webpage](https://nostarch.com/artofr.htm)
- *An Introduction to R for Spatial Analysis and Mapping* by Chris Brunsdon and Lex Comber, Sage, 2015. See book [webpage](https://uk.sagepub.com/en-gb/eur/an-introduction-to-r-for-spatial-analysis-and-mapping/book241031)
- *Geocomputation with R* by Robin Lovelace, Jakub Nowosad, Jannes Muenchow, CRC Press, 2019. See [online book](https://bookdown.org/robinlovelace/geocompr/).



## R

Created in 1992 by Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand

- Free, open-source implementation of *S*
    - statistical programming language 
    - Bell Labs


- Functional programming language
- Supports (and commonly used as) procedural (i.e., imperative) programming
- Object-oriented
- Interpreted (not compiled)



## Interpreting values

When values and operations are inputted in the *Console*, the interpreter returns the results of its interpretation of the expression

```{r, echo=TRUE}
2
"String value"
# comments are ignored
```



## Basic types

R provides three core data types

- numeric 
    - both integer and real numbers
- character 
    - i.e., text, also called *strings*
- logical
    - `TRUE` or `FALSE`


## Numeric operators

R provides a series of basic numeric operators



|Operator|Meaning         |Example  |Output     |
|--------|----------------|---------|-----------|
|+       |Plus            |`5 + 2`  |`r 5 + 2`  |
|-       |Minus           |`5 - 2`  |`r 5 - 2`  |
|`*`     |Product         |`5 * 2`  |`r 5 * 2`  |
|/       |Division        |`5 / 2`  |`r 5 / 2`  |
|%/%     |Integer division|`5 %/% 2`|`r 5 %/% 2`|
|%%      |Module          |`5 %% 2` |`r 5 %% 2` |
|^       |Power           |`5^2`    |`r 5^2`    |



```{r, echo=TRUE}
5 + 2
```



## Logical operators


R provides a series of basic logical operators to test



|Operator|Meaning           |Example           |Output              |
|--------|------------------|------------------|--------------------|
|==      |Equal             |`5 == 2`          |`r 5 == 2`          |
|!=      |Not equal         |`5 != 2`          |`r 5 != 2`          |
|> (>=)  |Greater (or equal)|`5 > 2`           |`r 5 > 2`           |
|< (<=)  |Less (or equal)   |`5 <= 2`          |`r 5 <= 2`          |
|!       |Not               |`!TRUE`           |`r !TRUE`           |
|&       |And               |`TRUE & FALSE`    |`r TRUE & FALSE`    |
|\|      |Or                |`TRUE | FALSE`    |`r TRUE | FALSE`    |



```{r, echo=TRUE}
5 >= 2
```



## Summary

An introduction to R

- Basic types
- Basic operators

**Next**: Core concepts

- Variables
- Functions
- Libraries
