# Core concepts{#core}

```{block2, type = 'rmdtip'}
In this lesson, we'll explore three fundamental concepts in programming:

1. Variables
2. Functions
3. Libraries

```

## Variables

Variables are essential for **storing data**. To define a variable, use an *identifier* (e.g., `a_variable`) on the left of an *assignment operator* `<-`, followed by the value you wish to assign (e.g., `1`):

```{r, echo=TRUE}
a_variable <- 1
```

To retrieve the value of the variable, simply use its **identifier**.

```{r, echo=TRUE}
a_variable
```

```{block2, type = 'rmdtip'}

To save your code, create an **R Script** in RStudio (File > New File > R Script). Select the code in the R Script Window and click 'Run' to execute it.

Note: Code execution is sequential, processing line by line.
```

Variables allow you to save the results of any computations performed in the code and retrieve them later for further analyses. For instance, you can declare a variable such as:

```{r, echo=TRUE}
a_variable <- 1
```

Then, you can manipulate the value of the variable as follows:

```{r, echo=TRUE}
a_variable <- a_variable + 10
```

Later in the code, you can assign the value to a different variable with:

```{r, echo=TRUE}
another_variable <- a_variable
```

At this point, you might wonder why bother using variables instead of simply typing the numbers directly. The answer is that variables make your code reusable and save you a lot of time.

**Let us consider the following example:**
  
Meteorologists monitor water temperature gradients in the Pacific Ocean to better understand El Niño weather patterns and forecast extreme weather conditions associated with it. In a given year, the water temperature at location A might be 22°C and at location B, 26°C. We could simply calculate the difference by executing the arithmetic operation '26 - 22' in the console window of RStudio. However, temperatures are measured in real-time, meaning we have to calculate temperature gradients repeatedly.

To expedite the process, we could write code that performs the calculation (temperature at location A minus temperature at location B). This piece of code uses two variables (temperatures at locations A and B) as inputs. As a result, we only need to update these two variables; the algorithm (simple subtraction in our example) is reusable.

Of course, the gains in efficiency are minor given the simplicity of this calculation. However, in more practical applications, the algorithm is likely to be composed of many lines of code that evaluate the risk of El Niño occurrence based on sensor records.

```{block2, type = 'rmdexercise'}

1) Create a new R script in RStudio (File > New File > R Script).
2) Declare two variables (`temp_A` and `temp_B`) and assign arbitrary temperature values to them.
3) Declare a third variable (`diff`) and assign the difference between the other two variables as its value.
4) Run your script (select your code and click 'Run').

<details closed>
<summary><ins>**See solution!**</ins></summary>
<p><i><font color="grey">
  
temp_A <- 24
temp_B <- 28
diff <- temp_A - temp_B

After executing the code in RStudio, you should notice changes in the Environment Panel, located in the top right corner. The **Environment Panel** will display three memory slots with identifiers `diff`, `temp_A`, and `temp_B`, having values of -4, 24, and 28, respectively. Invoking the name of an identifier in the code (e.g., typing `diff` and running it) will return the value stored in that memory slot.

To clear your workspace memory, click the broom icon in the Environment Panel's menu.

</font></i>
</p>
</details> 

```

## Algorithms and Functions

*"An algorithm is a mechanical rule, automatic method, or program for performing some mathematical operation"* (Cutland, 1980).

A *program* is a specific set of instructions that implements an abstract algorithm.

The definition of an algorithm (and thus a program) can consist of one or more functions. Functions are sets of instructions that perform a task, structuring code into functional, reusable units. Some functions receive values as inputs, while others return output values.

Programming languages typically provide pre-defined functions that implement common algorithms (e.g., finding the square root of a number or calculating a linear regression).

For example, the pre-defined function `sqrt()` calculates the square root of an input value. **Like all functions** in R, `sqrt()` is invoked by specifying the *function name* and *arguments* (input values) between parentheses:

```{r, echo=TRUE}
sqrt(2)
```

```{block2, type = 'rmdtip'}

Each input value corresponds to a parameter defined in the function. Sometimes, the parameter name must be specified. This concept will become clearer when you write your own functions later in the module.
   
```

`round()` is another predefined function in R:

```{r, echo=TRUE}
round(1.414214, digits = 2)
```

Note that the name of the second parameter (`digits`) needs to be specified. The `digits` parameter indicates the number of digits to keep after the decimal point.

The return value of a function can be stored in a variable:

```{r, echo=TRUE}
sqrt_of_two <- sqrt(2)
sqrt_of_two
```

Here, the output value is stored in a memory slot with the identifier `sqrt_of_two`. We can use the identifier `sqrt_of_two` as an argument in other functions:

```{r, echo=TRUE}
sqrt_of_two <- sqrt(2)
round(sqrt_of_two, digits = 3)
```

The first line calculates the square root of 2 and stores it in a variable named `sqrt_of_two`. The second line rounds the value stored in `sqrt_of_two` to three decimal places.

```{block2, type = 'rmdexercise'}

Can you store the output of the `round()` function in a second variable?

<details closed>
<summary><ins>**See solution!**</ins></summary>
<p><i><font color="grey">
  
sqrt_of_two <- sqrt(2)
rounded_sqrt_of_two <- round(sqrt_of_two, digits = 3)
    
</font></i>
</p>
</details> 

```

Functions can also be used as arguments within other functions. For instance, we can use `sqrt()` as the first argument in `round()`:

```{r, echo=TRUE}
round(sqrt(2), digits = 3)
```

In this case, the intermediate step of storing the square root of 2 in a variable is skipped.

```{block2, type = 'rmdtip'}
Using functions as arguments in other functions is often discouraged as it can make the code hard to understand.

Moreover, to enhance the readability of R code, it is recommended to follow naming conventions for variables and functions:
  
- R is a **case-sensitive** language:
    - UPPER and lower case are treated differently.
    - `a_variable` is different from `a_VARIABLE`.
- Names can include:
    - Alphanumeric characters.
    - `.` and `_`.
- Names must start with:
    - A letter.

```

## Libraries

Related, reusable functions in R can be grouped and stored in **libraries**, also known as *packages*.

As of now, there are more than 10,000 R libraries available. These can be downloaded and installed using the `install.packages()` function. Once a library is installed, the `library()` function is used to make it accessible in a script.

Libraries can vary greatly in size and complexity. For example:

- `base`: This includes basic R functions, such as the sqrt function discussed earlier.
- `sf`: A package providing [simple feature access](https://en.wikipedia.org/wiki/Simple_Features){target="_blank"}.

The `stringr` library illustrates the use of libraries in R. It offers a consistent and well-defined set of functions for string manipulation. Assuming the library is already installed on your computer, you can load it as follows:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(stringr)
```

If not yet installed, you can download and install it with:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
install.packages('stringr')  # Note: the function takes a string argument ('')
```

```{block2, type = 'rmdtip'}
Alternatively, libraries (a.k.a. packages) can be installed through RStudio's 'Install Packages' menu (Tools > Install Packages...). You can choose to install from CRAN or a package archive file. The majority of libraries are available through [CRAN - Comprehensive R Archive Network](https://cran.r-project.org/){target="_blank"}, a vast collection of R resources and libraries.

```

Once installed and loaded, the library provides a new set of functions in your environment. For instance, `str_length` in the `stringr` library returns the number of characters in a string:

```{r, echo=TRUE}
str_length("UNIGIS")
```

The `str_detect()` function returns `TRUE` if the first argument (a string) contains the second argument (a character string). Otherwise, it returns `FALSE`:

```{r, echo=TRUE}
str_detect("UNIGIS", "I")
```

`str_replace_all` replaces all instances of a specified character in a string with another character:

```{r, echo=TRUE}
str_replace_all("UNIGIS", "I", 'X')
```

```{block2, type = 'rmdtip'}

You can list all the functions available in the `stringr` library using the built-in function [`ls()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/ls){target="_blank"}:
```

```{r, echo=TRUE}
ls("package:stringr")
```