# Read and write data

In previous exercises we have **read** data from a CSV file into our script. Similarly we can also **write** code outputs to file.   

```{block2, type = 'rmdtip'}

In this lesson you will learn to read and write plain-text and spatial vector file formats. Moreover, we will retrieve online data by means of a data API.

```

## Read and write tabular data

A series of formats are based on plain-text files.

For instance...

- comma-separated values files `.csv`
- semi-colon-separated values files `.csv`
- tab-separated values files `.tsv`
- other formats using custom delimiters
- fix-width files `.fwf`

The [`readr` library](https://readr.tidyverse.org/){target="_blank"} (also part of [Tidyverse](https://www.tidyverse.org/packages/#import){target="_blank"}) provides a series of functions that can be used to load from and save to such data formats. For instance, the `read_csv` function reads a comma delimited (CSV) file from the path provided as the first argument. 

The code example below reads a CSV file that contains global fishery statistics provided by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators){target="_blank"} and queries Norwegian entries. The function `writes_csv`  writes these entries to a new CSV file.  

```{r, echo=TRUE, message=FALSE, warning = FALSE}

library(tidyverse)

fishery_data <- readr::read_csv("data/capture-fisheries-vs-aquaculture.csv")
#print(fishery_data)

#print(typeof(fishery_data$))

fishery_data %>%
  dplyr::filter(Entity == "Norway") %>%
  readr::write_csv("data/capture-fisheries-vs-aquaculture-norway.csv", append=FALSE) %>%
  
  dplyr::slice_head(n = 3) %>%
  knitr::kable()
  
```

In order to run the script, [download the CSV file](data/capture-fisheries-vs-aquaculture.csv){target="_blank"}. Then copy and run the code in a new R-script.

```{block2, type = 'rmdtip'}
Other important packages for reading tabular data are [readxl](https://readxl.tidyverse.org/){target="_blank"} for Excel (.xls and .xlsx) and [haven](https://haven.tidyverse.org/){target="_blank"} for SPSS, Stata and SAS data.  
```

## Read and write vector data

The library `sf` makes it easy to read and write vector datasets such as shapefiles. The name (sf stands for **simple features**) already implies that `sf` supports simple feature access via R.

Simple features is a widely supported data model that underlies data structures in many GIS applications including QGIS and PostGIS. A major advantage of this is that using the data model ensures your work is cross-transferable to other set-ups, for example importing from and exporting to spatial databases.

In order to load vector data in an R-Script, we can use the function `st_read()`. In the code block below, a [shapefile](data/nc.zip){target="_blank"} ([North Carolina sample data](https://r-spatial.github.io/sf/reference/nc.html){target="_blank"}) is loaded and assigned to a variable nc. 

The next line creates a basic map in `sf` by means of `plot()`. By default this creates a multi-panel plot, one sub-plot for each variable of the object. 

```{r, echo=TRUE, message=FALSE, warning = FALSE, results='hide'}

library(sf)

nc <- sf::st_read("data/nc.shp")

plot(nc)

```

The library `sf` represents features as records in a data frame or tibble with a geometry list-column. The example below renders three features (rows) of variable nc including the geometry column as well as the attributes `AREA` (feature area) and `NAME` (name of county):  

```{r, echo=TRUE, message=FALSE, warning = FALSE}

nc %>%
  dplyr::select(AREA, NAME, geometry) %>%
  dplyr::slice_head(n = 3) %>%
  knitr::kable()

```

`sf` also includes a number of operations to manipulate the geometry of features such as `st_simplify`:  

```{r, echo=TRUE, message=FALSE, warning = FALSE}

sf::st_simplify(nc) %>%
  plot(., max.plot = 1)

```

```{block2, type = 'rmdtip'}

You may have recognized that a dot (.) is used as a parameter in the function `plot()`. The dot represents the piped value. In the example above the dot is used to define the simplified geometry of nc as first parameter of function `plot()` and `max.plot = 1` as the second.  

```

In the next example, the `st_geometry()` retrieves the geometry attribute from variable nc, function `st_centroid()` calculates the centroid of the polygon geometry (counties) and function `st_write` writes the centroid point geometry to file. 

```{r, echo=TRUE, message=FALSE, warning = FALSE, results='hide'}

sf::st_geometry(nc) %>%
  sf::st_centroid() %>%
  sf::st_write("data/nc-centroids.shp", delete_dsn = TRUE) %>%
  plot(pch = 3, col = 'red') 
  
```

```{block2, type = 'rmdtip'}
The online book [Geocomputation with R](https://r.geocompx.org/){target="_blank"} offers a more comprehensive explanation of available geometric, attribute and spatial data operations. For a quick overview, you may turn to the [sf cheatsheets](https://github.com/rstudio/cheatsheets/blob/master/sf.pdf){target="_blank"}. 
```

In order to test the code on your machine, [download](data/nc-centroids.zip){target="_blank"} the North Carolina dataset and install the libraries `sf` and `Rcpp` before you run the code in an R-Script.   

```{block2, type = 'rmdexercise'}

The `plot()` function offers a large number of arguments that can be used to customize your map. Replace 'Area' in the map above by a more meaningful map title. Turn to the [documentation](https://r-spatial.github.io/sf/reference/plot.html){target="_blank"} for more information.    

See [my solution](Solution_Miniexercise_plot.R){target="_blank"}!

```

```{block2, type = 'rmdtip'}
Similar R functions are also available for raster data (see package [raster: Geographic Data Analysis and Modeling](https://cran.r-project.org/web/packages/raster/raster.pdf){target="_blank"}) 
```

## Data API

> API is the acronym for Application Programming Interface, which is a software intermediary that allows two applications to talk to each other.

By means of an API we can read, write and modify information on the web. The following video briefly introduces the technology behind it.

```{r APIvideo, echo=FALSE, fig.align='center', fig.cap="Video (3:13 min): REST API concepts and examples."}

knitr::include_url("https://www.youtube.com/embed/7YcW25PHnAA?end=194")

``` 

The most important take away messages are:

1) With a REST API web data is accessible through a URL (Client-Server call via HTTP protocol)
2) The HTTP **Get** Method delivers data (a Response) - i.e. is used to read data, the HTTP **Post** Method is used to create new REST API resources (write data).
3) URL Parameters are used to filter specific data from a response.
4) Typically, APIs can return data in a number of different formats. 
5) JSON is a very popular format for transferring web data.
6) The two primary elements that make up JSON are keys and values.

The [library `httr`](https://cran.r-project.org/web/packages/httr/vignettes/quickstart.html){target="_blank"} (also part of [Tidyverse](https://www.tidyverse.org/packages/#import){target="_blank"}) offers functions to programmatically implement API calls in an R script. We will make use of this library to let our R script interact with the [APIs](https://openweathermap.org/api){target="_blank"} offered by [OpenWeather Map](https://openweathermap.org/){target="_blank"} that contain historical and real-time weather data for retrieval. In the upcoming example we will make a call to the [Current Weather API](https://openweathermap.org/current){target="_blank"}, which is one out of currently [10 free APIs](https://openweathermap.org/price){target="_blank"} provided by OpenWeather Map.

For accessing the data, we need to create a URL that is composed of a reference to the data source (base) and parameters to filter the desired data subset (lat and lon). The parameters are concatenated by means of a '&' symbol, a question mark '?' needs to be placed between reference URL and parameters: 

```{r, echo=TRUE, message=FALSE, warning = FALSE}

library(httr)

base <- "http://api.openweathermap.org/data/2.5/weather"
lat <- "47.81"    
lon <- "13.03"
apiKey <- "3f87141421b32590d50416aae5ca780c"
  
call <- paste(base, "?lat=", lat, "&lon=" , lon, "&appid=", apiKey, sep="")

print(call)
  
get <- GET(call)

```

In the code above, the function `paste()` assembles base URL and parameters. The URL (of type string) is then passed as an argument to function `GET()` that executes the HTTP Get method. 

```{block2, type = 'rmdtip'}
Note: The OpenWeather API requires an API Key to be passed as a parameter in the call. Get your [personal Key](https://openweathermap.org/appid){target="_blank"} to implement your own API requests. 

If your code returns an error 401, this most likely indicates that your key is not activated yet. According to the [FAQs](https://openweathermap.org/faq){target="_blank"}, it may take a couple of hours until your key is active. 

```

Per default, the `GET()` function returns a response object. Printing a response object gives you some useful information: the actual url used (after any redirects), the http status, the file (content) type, the size, and if it’s a text file, the first few lines of output.

You can pull out important parts of the response with various helper functions such as `status_code()` and `content()`:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

status_code(get)

str(content(get))

```

The Current Weather API call returned a number of real-time weather variables such as temperature, air pressure or humidity for the location of Salzburg. Current weather data may alternatively be retrieved by City ID or City Name (see the [documentation](https://openweathermap.org/current){target="_blank"} to get an overview of available API parameters).

```{block2, type = 'rmdexercise'}

Inspect the response object by means of a function called `headers()`. What methods are allowed when accessing the Current Weather API?

<details closed>
<summary><ins>**See solution!**</ins></summary>
<p><i><font color="grey">
  
Only GET and POST methods are allowed.

Other APIs allow to update existing REST API resources (PUT method) or to delete a REST API resource (DELETE method). 
    
</font></i>
</p>
</details>   
  
```

In order to facilitate subsequent analyses and data visualization, we can convert the content of the return object to a `data frame` by means of function `fromJSON()` that is part of the library `library(jsonlite)`:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

library(jsonlite)

get_text <- content(get, "text")                 #retrieve contents of request as character vector (library httr)
get_json <- fromJSON(get_text, flatten = TRUE)   #convert from JSON to R Object (library jsonlite)
get_df <- as.data.frame(get_json)                #convert R Object to Data Frame 

```

```{block2, type = 'rmdexercise'}
Copy the code snippets above to a new R-Script. **Make sure to replace the key in the code example by your own API key!** If you need help, please turn to the discussion forum.
```

```{block2, type = 'rmdtip'}

[Tidyverse](https://www.tidyverse.org/packages/){target="_blank"} also provides other packages for reading data such as [DBI](https://cran.r-project.org/web/packages/DBI/){target="_blank"} for relational databases [jsonlite](https://cran.r-project.org/web/packages/jsonlite/){target="_blank"} for JSON and [xml2](https://cran.r-project.org/web/packages/xml2/){target="_blank"} for XML. 

```


