# Spatial Data Manipulation

## Introduction
- Briefly discuss the importance of spatial data wrangling in geoinformatics and related fields.
- Quick highlight of the difference between traditional data wrangling and spatial data wrangling.

## Spatial Aggregation
- Aggregating spatial data based on certain attributes or spatial units.
- Introduce the sf and terra packages for aggregation tasks.

```{r}
# Example code for spatial aggregation using sf and terra
```



## Merging and Appending Spatial Data
- Combining multiple spatial datasets.
- Introduce the sf package for these tasks.
```{r}
# Example code for merging and appending using sf
```

## Handling Missing Data
- Discuss methods to identify, visualize, and impute missing data.
- Introduce the **tidyverse** suite for handling missing data in spatial datasets. (but again, watchout for **httr1** and **httr2** + (**time of writing 12.09.2023**)

```{r}
# Example code for handling missing data using tidyverse
```

## Spatial Data Cleaning
- Detecting and correcting topological errors.
- Introduce the sf and spatstat packages for these tasks.

```{r}
# Example code for spatial data cleaning using sf and spatstat
```

## Multidimensional Data Wrangling with stars
- Handling and manipulating multidimensional arrays.
- Discuss the importance of stars in the R spatial ecosystem.

```{r}
# Example code for wrangling multidimensional data using stars
```

## Challenges in Spatial Data Structures
- Issues with data integrity and accuracy.
- Handling different coordinate reference systems (CRS). (If you get for example a old or exotic dataset)

## Dealing with large spatial datasets.
- Dive Deeper: Strategies for optimizing spatial data structures for large datasets and high-performance computing.

## Case Study: A Complete Spatial Data Wrangling Workflow
- Would be a quite interesting change of wind: a comprehensive case study integrating various spatial data wrangling techniques...
- This could be based on a real-world scenario, such as urban planning, environmental monitoring, or transportation.
- Alternatively, a set of exercises allowing our students to apply the techniques learned in the chapter.


---


## Introduction
- Briefly explain the difference between data wrangling and data manipulation in the spatial context.
- Highlight the importance of spatial data manipulation in GIS and spatial analysis workflows.
- Introduce the primary packages for this chapter: sf, terra, stars, and spatstat.

## Filtering Spatial Data
- Filtering sf Objects
- Demonstrate how to filter spatial features in sf objects based on attribute conditions.
- Introduce the sf package for vector data filtering.
```{r}
# Example code for filtering sf objects
```

## Filtering Rasters with terra
- Show how to filter raster data based on cell values or spatial conditions using terra.
```{r}
# Example code for filtering rasters with terra
```
## Spatial Joins with sf (though.. we could also put that in Chapter 10 and make 12 bigger with spatial stat)
- Explain the concept of spatial joins.
- Demonstrate how to join two spatial datasets based on their spatial relationship using sf.
```{r}
# Example code for spatial joins with sf
```
## Geometric Operations with sf
### Intersect
- Demonstrate how to find the common spatial elements between two datasets using sf.
```{r}
# Example code for intersect operation with sf
```
### Buffer
- Show how to create a buffer zone around a spatial feature using sf.
```{r}
# Example code for buffer operation with sf
```
### Union and Difference
- Explain and demonstrate the union and difference operations between two spatial datasets using sf.
```{r}
# Example code for union and difference operations with sf
```
##  Coordinate Transformations and Reprojections with sf and terra
- Discuss the importance of having datasets in the same coordinate system.
- Demonstrate how to reproject spatial data using both sf for vector data and terra for raster data.
```{r}
# Example code for coordinate transformations with sf and terra
```
## Raster Operations with terra
### Resampling
- Show how to change the resolution of raster datasets using terra.
```{r}
# Example code for raster resampling with terra
```
### Raster Algebra
- Demonstrate basic raster algebra operations like addition, subtraction, and multiplication of raster layers using terra.
```{r}
# Example code for raster algebra with terra
```
## Applied Spatial Statistics with spatstat 
(I think, this will be yours Christian :) But it might be a pretty good idea to showcase it in chapter 12, atleast basics.):
- Introduction to spatstat:
- Briefly introduce the spatstat package and its capabilities in handling advanced spatial statistics.
- Point Pattern Analysis (ppa):
- Dive deeper into point pattern analysis, discussing its importance and applications in spatial statistics.
- Spatial Clustering:
- Discuss the concept of spatial clustering and its relevance in identifying patterns in spatial data.
- Illustrate with examples how spatstat can be used for spatial clustering analysis.
- Nearest Neighbor Analysis:
- Introduce Nearest Neighbor as a crucial type of spatial query often used in spatial statistics to analyze the spatial distribution of features.
- Discuss its applications in evaluating the spatial arrangement of points, assessing spatial randomness, and identifying - patterns such as clustering or dispersion.
- Provide practical examples demonstrating how to perform Nearest Neighbor analysis using spatstat, and discuss the interpretation of the results.


## Comparative Analysis of Packages
- After that intense usage of packages, Investigate and compare the performance, ease of use, and capabilities of sf, terra, stars, and spatstat for various spatial operations. Compiled languages (for example via C, C++) and packages and their difference in speed.

## Exercise
- Provide exercises that allow readers to apply the techniques learned in the chapter, ensuring they get hands-on experience with each of the introduced packages.