# Spatial Data Manipulation {#spm}

In the previous lesson, you learned how to access data on a server via a REST API. The Open Geospatial Consortium (OGC) has adapted the REST API paradigm for geospatial applications. A variety of OGC API standards have been implemented to provide and utilize geospatial data on the web. An overview of the current implementation status can be found [here](https://ogcapi.ogc.org/#standards){target="_blank"}.

```{block2, type = 'rmdtip'}

In this lesson, we will utilize online data resources in workflows that involve data cleaning, spatial queries, and analyses. Additionally, you will learn to carry out basic raster manipulation operations on a sample raster dataset.

```

## Data Acquisition

We will work with vector data to correctly identify agricultural land parcels in European Union countries. The Austrian Agricultural Agency ([AMA](https://www.ama.at/intro){target="_blank"}) provides access to Austrian agricultural parcels through an [OGC Rest API - Feature](https://ogcapi.ogc.org/){target="_blank"} interface. The OGC API Features standard is the successor to the OGC Web Feature Service (WFS) specification.

```{block2, type = 'rmdtip'}

The R syntax used to interact with OGC APIs is the same as described in the "Data API" section (see Lesson [8](#readwrite){target="_blank"}).

```

Before loading the data into an R script, examine the API's contents. Visit the web service's [landing page](https://gis.lfrz.gv.at/api/geodata/i009501/ogc/features/v1/){target="_blank"} and proceed to the [collection page](https://gis.lfrz.gv.at/api/geodata/i009501/ogc/features/v1/collections?f=text%2Fhtml){target="_blank"}. There you will find an overview of the available layers.

Enter the following URL into your browser:

```{}
https://gis.lfrz.gv.at/api/geodata/i009501/ogc/features/v1/collections/i009501:invekos_feldstuecke_2024_1/items?f=json&limit=10
```

This request returns the first ten features of the `i009501:invekos_feldstuecke_2024_1` layer (agricultural land parcels as polygons) in GeoJSON format.


> GeoJSON is a JSON-based standard for representing simple geographic features, along with their non-spatial attributes. 

Upon inspecting the GeoJSON, you will find the coordinate vertices of the polygon features and attributes such as `fs_flaeche_ha` (parcel area in hectares) and `fnar_bezeichnung` (land use).


Use the `bbox` parameter to filter resources within a specific area:

```{}
https://gis.lfrz.gv.at/api/geodata/i009501/ogc/features/v1/collections/i009501:invekos_feldstuecke_2024_1/items?f=json&bbox=14,48,14.02,48.02
```

For a visual representation of the bounding box, enter the coordinates `14,48,14.02,48.02` into  [linestrings.com](https://linestrings.com/bbox/){target="_blank"} and click "Display box".

To execute the request in R, use the following script:

```{r, echo=TRUE, message=FALSE, warning = FALSE}
library(httr2)
library(geojsonsf)
library(sf)

# Define the new API endpoint
full_url <- "https://gis.lfrz.gv.at/api/geodata/i009501/ogc/features/v1/collections/i009501:invekos_feldstuecke_2024_1/items?f=json&bbox=14,48,14.02,48.02"

# Make the request and convert the response to an sf object
invekos <- request(full_url) %>%   # Create request
  req_perform() %>%                # Execute request
  resp_body_string() %>%           # Extract JSON body as string
  geojsonsf::geojson_sf()          # Convert JSON string to sf object

# View the first few rows of the data
print(head(invekos))

```

The above code retrieves polygon features within the specified bounding box and stores them in an object named `invekos`.

## Data Cleaning

Once the data is loaded into R, we can more closely examine its structure. The `dplyr` function `glimpse`, for instance, allows us to preview each column's name and type:

```{r, echo=TRUE, message=FALSE, warning = FALSE}
library(dplyr)
dplyr::glimpse(invekos)

```

> The abbreviation `dbl` stands for `double`, a data type that stores numbers with decimal points.

To create a subset of the `invekos` sf-object, use the following code:

```{r, echo=TRUE, message=FALSE, warning = FALSE}
invekos %>%
  head(2) %>%
  dplyr::select(fnar_bezeichnung, fs_flaeche_ha, geo_id, geometry) %>%
  knitr::kable(., format="html")

```

Field names and entries are in German. To rename them, use the base R function `colnames()`:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

# Subset of invekos object
invekos.sub <- invekos %>%
  dplyr::select(fnar_bezeichnung, fs_flaeche_ha, geo_id, geometry)

# Renaming fields
colnames(invekos.sub)[1] <- "land_use"
colnames(invekos.sub)[2] <- "area_ha"

```

Entries are renamed using `dplyr`'s `mutate` and `case_when` functions:

```{r, echo=TRUE, message=FALSE, warning = FALSE}
unique(invekos.sub$land_use)  # Unique entries in 'land_use'

# Renaming entries 'ACKERLAND' and 'GRÜNLAND'
invekos.sub <- dplyr::mutate(invekos.sub, land_use = 
                               case_when(land_use == 'ACKERLAND' ~ 'arable land',
                               land_use == 'GRÜNLAND' ~ 'grassland', TRUE ~ 'Other'))

# Display result
invekos.sub %>%
  head(2) %>%
  knitr::kable(., format="html")

```

For an initial visual impression, plot the sf-object using the base R `plot()` function:

```{r, echo=TRUE, message=FALSE, warning = FALSE}
plot(invekos.sub[1], main="Land Use", key.pos = 1, key.width = lcm(1.8))
```

By default, the `plot()` function generates a multi-panel plot, with one sub-plot for each field of the object. However, specifying `invekos.sub[1]` restricts the output to a single plot, specifically showcasing the `land_use` field. The function parameter `key.pos = 1` aligns the legend below the map (1=below, 2=left, 3=above and 4=right). `key.width` defines the width of the legend. 

Next, we can validate the geometry of sf polygons using `st_is_valid`:

```{r, echo=TRUE, message=FALSE, warning = FALSE}
sf::st_is_valid(invekos) %>%
  summary()
```

The `st_is_valid` function checks the validity of each geometry, returning a logical vector. The summary confirms that all geometries in our dataset are valid. If any invalid geometries are detected, they can be corrected using the `st_make_valid` function.

```{block2, type = 'rmdexercise'}

Let's closely investigate a polygon geometry with the code below, which creates a single polygon feature. The validity check flags an invalid geometry. Can you identify the problem?

```

```{r, echo=TRUE, message=FALSE, warning = FALSE}
# Coordinates for the polygon
coords <- matrix(c(-1, -1, 1, -1, 1, 1, 0, -1, -1, -1), ncol = 2, byrow = TRUE)

# List of coordinate matrices
list_of_coords <- list(coords)

# Constructing the polygon
polygon <- st_polygon(list_of_coords)

# Converting to an sf object
error_sf <- st_sf(geometry = st_sfc(polygon))

# Validity check
sf::st_is_valid(error_sf)

```

<details closed>
<summary><ins>**See solution!**</ins></summary>
<p><font color="grey">
  
When plotted, the polygon reveals a [sliver polygon](https://en.wikipedia.org/wiki/Sliver_polygon){target="_blank"}, which is invalid due to its shape.

```{r, echo=TRUE, message=FALSE, warning = FALSE}

plot(error_sf)

```

Another common cause of invalid geometries is self-intersection of lines.

> Note: A valid polygon requires at least four coordinate points with the first and last points being identical. A polygon with only three points is considered to have an invalid geometry.

</font>
</p>
</details>

## Vector operations

The structure of an `sf` object is similar to that of data frames, enabling attribute-based filtering operations using `dplyr` functions, as described in Lesson [7](#datman){target="_blank"}.

For example, to extract grassland parcels, the `filter()` function is used:

```{r, echo=TRUE, message=FALSE, warning = FALSE}
invekos.sub %>%
  dplyr::filter(land_use=='grassland') %>%
  {plot(.[1], main="Land Use", key.pos = 1, key.width = lcm(1.8))}
```

> Note the use of curly brackets in the `plot()` function. This is required because the dot notation (.) cannot directly index a piped value. For a deeper explanation, see this [Stack Overflow discussion](https://stackoverflow.com/questions/42385010/using-the-pipe-and-dot-notation){target="_blank"}.

### Geometrical operations

The `sf` package provides a range of geometric predicates such as `st_within`, `st_contains`, or `st_crosses`. A complete list is available [here](https://r-spatial.github.io/sf/reference/geos_binary_pred.html){target="_blank"}. While these functions are often used between pairs of simple feature geometry sets, they can also operate on a single `sf` object:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

sf::st_intersects(invekos.sub, sparse = TRUE)

```

This function returns a sparse matrix revealing intersections within features of the same `sf` object `invekos.sub`, indicating possible topology errors. For correcting these errors, GIS software like [QGIS](https://www.qgis.org/de/site/){target="_blank"} is recommended as R's capabilities for handling topology errors are less advanced.

### Binary operations

Binary operations are essential for analyzing spatial relationships between different datasets. Here, we demonstrate this by creating an `sf` object representing farmsteads using data from the [AMA Rest API service](https://gis.lfrz.gv.at/api/geodata/i009501/ogc/features/v1/collections?f=text%2Fhtml){target="_blank"}:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

# Define the new API endpoint for farmsteads
farms_url <- "https://gis.lfrz.gv.at/api/geodata/i009501/ogc/features/v1/collections/i009501:invekos_hofstellen_2024_1/items?f=json&bbox=14,48,14.02,48.02&limit=100"

# Make the request and convert the response to an sf object
farms <- httr2::request(farms_url) %>%   # Create request
  httr2::req_perform() %>%               # Execute request
  httr2::resp_body_string() %>%          # Extract JSON body as string
  geojsonsf::geojson_sf()                # Convert JSON string to sf object

# Plot land parcels with farm points overlayed
plot(invekos.sub[1], main = NULL, key.pos = NULL, reset = FALSE, col_graticule = "grey")
plot(farms[1], main = NULL, key.pos = NULL, pch = 7, col = 'red', add = TRUE, cex = 1)

```

> Note: The `plot()` function is used sequentially to overlay farm points on top of the land parcel polygons. For advanced plotting techniques, refer to the [`plot()` documentation](https://r-spatial.github.io/sf/reference/plot.html){target="_blank"}.

We proceed to calculate the distances between farms and land parcels to determine proximity:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

dist_m <- sf::st_distance(farms, invekos.sub)

```

The `st_distance` function calculates the shortest distance matrix between the geometries of the two `sf` objects, with distances returned in meters.

```{block2, type = 'rmdexercise'}

Distances are computed in meters, which may seem unexpected since the reference system's units are in degrees. To understand the underlying calculations, examine the coordinate reference system of the `sf` objects `farms` and `invekos.sub` using `sf::st_crs()`. This will reveal that objects use a projected coordinate system (EPSG:31287), which allows for metric distance calculations.

In this short exercise, we will take a closer look at the algorithm that is implemented in the function `st_distance`.

Open the [documentation of `st_distance`](https://r-spatial.github.io/sf/reference/geos_measures.html){target="_blank"} to find out how metric distances are derived from geographic coordinates.

<details closed>
<summary><ins>**See solution!**</ins></summary>
<p><font color="grey">

According to the documentation, metric distances are computed based on the Coordinate Reference System (CRS) of the `sf` objects. Since both `farms` and `invekos.sub` use **EPSG:31287** (a projected CRS), the `st_distance` function calculates planar (Euclidean) distances in meters. This is possible because projected CRSs preserve distance relationships locally, unlike geographic CRSs which use degrees.

For more detailed information on distance calculations, refer to [Algorithms for geodesics, Journal of Geodesy](https://link.springer.com/article/10.1007/s00190-012-0578-z){target="_blank"}.

</font>
</p>
</details>

```

When plotting the complete distance matrix `dist_m`, we see that column 1 contains the distances between the first feature (parcel 1) in the `invekos.sub` `sf` object and the 100 farm features of the farms sf object. Accordingly, the matrix has 100 columns (one for every parcel) and 100 rows (one for every farm).

The following line returns the first column of the distance matrix as a vector:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

dist_m[, 1]

```

To identify the farm that is located closest to parcel 1, we need to query the index of the minimum value in this vector:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

which(dist_m[, 1] == min(dist_m[, 1]))


```

The `which` function, a base R utility, identifies the indices of elements that satisfy a given condition. In the example, it returns the index of the smallest value within the vector `dist_m[, 1]`. Consequently, it indicates which farm is the closest to parcel 1.

The demonstrated procedure for detecting closest farms can be executed for every parcel in a for-loop. The number of the closest farm is appended to a vector named `closest`. This vector is in turn appended as a new column to the `sf` object `invekos.sub`. Finally, `invekos.sub` is plotted together with `farms`:

There are now two approaches to perform this loop:

**Approach 1: Dynamic Loop Using `ncol(dist_m)`**

This method dynamically adjusts the loop to the number of parcels by using `ncol(dist_m)`. It's flexible and scalable, making it suitable for datasets where the number of parcels may change.

```{r, echo=TRUE, message=FALSE, warning = FALSE}


# Approach 1: Dynamic Loop Using ncol(dist_m)
closest <- c()

for (i in 1:ncol(dist_m)){
  
  out <- which(dist_m[, i] == min(dist_m[, i]))
  
  closest <- c(closest, out)
  
}

cbind(invekos.sub, closest) %>%
  {plot(.[4], main = NULL, key.pos = NULL, reset = FALSE)}

plot(farms[1], main = NULL, key.pos = NULL, pch = 7, col = 'red', add = TRUE, cex = 1)


```

**Approach 2: Fixed Loop Using a Hardcoded Value (e.g., 100)**

This method uses a fixed number of iterations (100) based on prior knowledge of the dataset size. It's simpler but less flexible, as it requires manual adjustment if the number of parcels changes.

```{r, echo=TRUE, message=FALSE, warning = FALSE}


# Approach 2: Fixed Loop Using a Hardcoded Value (e.g., 100)
closest <- c()

for (i in 1:100){
  
  out <- which(dist_m[, i] == min(dist_m[, i]))
  
  closest <- c(closest, out)
  
}

cbind(invekos.sub, closest) %>%
  {plot(.[4], main = NULL, key.pos = NULL, reset = FALSE)}

plot(farms[1], main = NULL, key.pos = NULL, pch = 7, col = 'red', add = TRUE, cex = 1)


```

The output map visualizes the closest farm to each agricultural parcel, highlighting the practical application of `sf` Geometrical operations.

This chapter focuses on logical matrix outputs from geometric operations. For operations generating new geometries, such as `st_union`, `st_buffer`, and `st_centroid`, see [operations returning a geometry](https://r-spatial.github.io/sf/articles/sf3.html#operations-returning-a-geometry){target="_blank"}. For network operations on `sf` objects, consider using `sfnetworks` or the `igraph` package.

## Raster operations

Having previously discussed the structure of `SpatRaster Objects` in Lesson [4](#spds){target="_blank"} and the reading and writing of such objects in Lesson [8](#readwrite){target="_blank"}, we now turn our attention to raster manipulation operations like *resampling* and *cropping*.

> For the following examples, we will utilize a sample dataset from the `terra` package. You can download the sample data [here](data/terra_sample.tif){target="_blank"}).  

### Resampling

Resampling is crucial for working with raster datasets. It alters the spatial resolution, allowing you to align multiple rasters with different resolutions. Additionally, it adjusts the level of detail necessary for your analysis.

Let's demonstrate resampling with the `terra` sample dataset:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

library(terra)

r <- terra::rast("data/terra_sample.tif") # path may be different on your machine
plot(r, main='SpatRaster from file')

```

Before we change the raster resolution of `SpatRaster Object`, it is important to know the original resolution of the raster. You can use the `res()` function to check the original resolution:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

res(r)

```

For resampling, we'll create a target raster by copying the original and setting a new resolution:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

r2 <- r
terra::res(r2) <- 80 # Assigning a new resolution of 80x80

```

Although this operation clears the raster values, `r2` can still be used as a target for resampling:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

r_resampled <- terra::resample(r, r2, method="bilinear")

res(r2) # Verify the new resolution
plot(r_resampled, main='Resampled 80x80')

```

The `bilinear` method is used for interpolating new values in `r_resampled`, with alternatives such as `nearest` or `cubic`. The choice of interpolation method for resampling raster data is a crucial consideration, discussed in further detail [here](https://gis.stackexchange.com/questions/2587/deciding-what-interpolation-method-to-use-for-resampling-raster-data){target="_blank"}. 

### Crop raster

Raster cropping allows you to select a specific area from a larger raster dataset for targeted analysis. Both `SpatRaster` and `SpatExtent` objects can be utilized for cropping operations.

To derive an extent from a `SpatRaster`, the `ext` function is used:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

terra::ext(r)

```

The output format is `<xmin, xmax, ymin, ymax>`. Knowing this, we can define a cropping extent and apply it using the `crop` function:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

crop_ext <- terra::ext(180000, 180200, 331000, 331250)
subset_r <- terra::crop(r, crop_ext)
plot(subset_r, main='Subset of r')

```

If the resulting cropped area doesn't match expectations due to cell alignment, adjust the cropping extent or raster resolution as necessary.

### Raster algebra

`Terra` provides functionality for algebraic operations on rasters, supporting standard arithmetic and logical operators, and functions like `abs` and `round`. For a full list, visit the [terra algebra documentation](https://rspatial.org/pkg/4-algebra.html){target="_blank"}.

Here's how you can perform some simple operations:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

# Add a constant value to each cell
s <- r + 10
# Calculate the square root of cell values
s <- sqrt(r)
# Generate a logical raster based on cell value equality
s1 <- s == 15

plot(s1)

```

Operators can be applied between `SpatRaster` objects with matching resolutions and origins. The result covers their spatial intersection.

You can also perform in-place replacements:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

r[r == 255] <- 2550
plot(r)

```

Here, all cells with the value `255` are replaced with `2550`.

A broader range of vector and raster operations can be found in [Chapter 5 - Geometry Operations, in the Book Geocomputation with R](https://bookdown.org/robinlovelace/geocompr/geometric-operations.html){target="_blank"}.

```{block2, type = 'rmdtip'}
**Note:** As of writing this module, "Geocomputation with R" uses the `raster` package, which is expected to be superseded by `terra`.
```