# Spatial Data Manipulation

In the previous lesson you have learned to get access to data on a server via Rest API ... few words about OGC Rest API

In this lesson we will....

## Acquire data

Invekos Service Metadata: https://www.data.gv.at/katalog/dataset/56d93431-af47-4fc0-a014-012759d8db4e#applications

OGC API - Feature: https://gis.lfrz.gv.at/ogcapi009501/ogc/features/

Collection Page: https://gis.lfrz.gv.at/ogcapi009501/ogc/features/collections?f=text%2Fhtml

```{}

#get 10 features of INVEKOS feldstuecke
#copy to browser
https://gis.lfrz.gv.at/ogcapi009501/ogc/features/collections/ogcapi009501:INVEKOS_feldstuecke_aktuell_polygon/items?f=json&limit=10

```

```{}

# request with bbox
https://gis.lfrz.gv.at/ogcapi009501/ogc/features/collections/ogcapi009501:INVEKOS_feldstuecke_aktuell_polygon/items?f=json&bbox=14,48,14.02,48.02 

```

enter coordinates to display bounding box: https://linestrings.com/bbox/

```{r, echo=TRUE, message=FALSE, warning = FALSE}

library(httr2)
library(geojsonsf)
library(sf)

full_url <- "https://gis.lfrz.gv.at/ogcapi009501/ogc/features/collections/ogcapi009501:INVEKOS_feldstuecke_aktuell_polygon/items?f=json&bbox=14,48,14.02,48.02"

invekos <- httr2::request(full_url) %>%   #Create request
  httr2::req_perform() %>%           #Execute request
  httr2::resp_body_string() %>%      #Extract JSON body as string
  geojsonsf::geojson_sf()            #JSON string to sf object

```

## Data Cleaning

select columns and show polygon features 1 and 2:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

invekos %>%
  head(2) %>%
  dplyr::select(fnar_bezeichnung, fs_flaeche_ha, geo_id, geometry) %>%
  knitr::kable(., format="html")

```

`glimpse` may be used to see every column name and type, dbl stands for double, which is a numeric data type

```{r, echo=TRUE, message=FALSE, warning = FALSE}

dplyr::glimpse(invekos)

```

As a next step we can validate the geometry of sf polygons. Four point coordinates are required to create a closed polygon shape (*Remember:* First and last point need to be identical). A polygon with only three points is classified as having an invalid geometry. Sf-Function `st_is_valid` is used to detect invalid geometries:

```{r, echo=TRUE, message=FALSE, warning = FALSE}

sf::st_is_valid(invekos)
summary(invekos)

```

The function returns a logical vector indicating for each geometry whether it is valid. The summary statistic shows that the vector is composed of 100 TRUE values. Accordingly, our geometries are valid. Given that function `st_is_valid` returns FALSE, `st_make_valid` can be used to return objects with topologically correct geometry.

```{block2, type = 'rmdexercise'}

The code below creates a single polygon feature. However, the validity check indicates an invalid geometry. What is problem with this polygon?

<details closed>
<summary><ins>**See solution!**</ins></summary>
<p><font color="grey">
  
Plotting the polygon (`plot(error_sf)`) shows a [sliver polygon](https://en.wikipedia.org/wiki/Sliver_polygon){target="_blank"}. Another reason for a non-valid geometry may be the self-intersection of lines. 
  
</font>
</p>
</details>

```

```{r, echo=TRUE, message=FALSE, warning = FALSE}
# Create a matrix of coordinates
coords <- matrix(c(-1,-1, 1,-1, 1,1, 0,-1, -1, -1), ncol = 2, byrow = TRUE)

# Create a list of matrices (in this case, just one matrix)
list_of_coords <- list(coords)

# Create the polygon
polygon <- st_polygon(list_of_coords)

# Convert to spatial feature
error_sf <- st_sf(geometry = st_sfc(polygon))

sf::st_is_valid(error_sf)

```



german attribute names and notations (e.g. Ackerland) to english
filter important information (e.g. fnar_bezeichnung, fs_flaeche_ha, geo_id, geometry), see above
check topology

## Spatial filtering

## Spatial joins

## Geometrical operations

See geometrical operations:
https://r-spatial.github.io/sf/articles/sf3.html#geometrical-operations

Union, buffer, intersect etc.
