# New Chapter: Spatial Data Wrangling
## Introduction
- Briefly discuss the importance of spatial data wrangling in geoinformatics and related fields.
- Quick highlight of the difference between traditional data wrangling and spatial data wrangling.

## Spatial Aggregation
- Aggregating spatial data based on certain attributes or spatial units.
- Introduce the sf and terra packages for aggregation tasks.

```{r}
# Example code for spatial aggregation using sf and terra
```

## Coordinate Transformations
- See Chapter 10 prototype
- Changing the spatial reference system of datasets.
- Discuss the importance of coordinate transformations in spatial analysis.
- Introduce the sf package for these tasks.
- Visualizing the results of coordinate transformations.
- Then, for example Introduce the ggspatial package for visualizing transformed spatial data in conjunction with ggplot2. (**Note that, ggspatial was introduced in chapter 8. Data Visualisation.. Very briefly.**)

```{r}
# Example code for coordinate transformations using sf
```

## Raster Resampling and Reprojection
- Adjusting the resolution of raster datasets or changing their projection.
- Introduce the terra package for these tasks.
```{r}
# Example code for raster resampling and reprojection using terra
```

## Merging and Appending Spatial Data
- Combining multiple spatial datasets.
- Introduce the sf package for these tasks.
```{r}
# Example code for merging and appending using sf
```

## Handling Missing Data
- Discuss methods to identify, visualize, and impute missing data.
- Introduce the tidyverse suite for handling missing data in spatial datasets. (but again, watchout for httr1 and httr2 + (**time of writing 12.09.2023**)

```{r}
# Example code for handling missing data using tidyverse
```

## Spatial Data Cleaning
- Detecting and correcting topological errors.
- Introduce the sf and spatstat packages for these tasks.

```{r}
# Example code for spatial data cleaning using sf and spatstat
```

## Multidimensional Data Wrangling with stars
- Handling and manipulating multidimensional arrays.
- Discuss the importance of stars in the R spatial ecosystem.

```{r}
# Example code for wrangling multidimensional data using stars
```

## Case Study: A Complete Spatial Data Wrangling Workflow
- Would be a quite interesting change of wind: a comprehensive case study integrating various spatial data wrangling techniques...
- This could be based on a real-world scenario, such as urban planning, environmental monitoring, or transportation.
- Alternatively, a set of exercises allowing our students to apply the techniques learned in the chapter.
