# New Chapter: **Spatial Data Wrangling (see modified chapters)**
## Spatial Aggregation
First: Aggregating spatial data based on certain attributes or spatial units. For instance, summing up population data by for example administrative boundaries.

```{r}
# Example code for spatial aggregation
```

## Coordinate Transformations
Changing the spatial reference system of datasets to ensure they're on the same projection or to transform them to a more suitable projection for analysis. (needs to be checked, if it makes sense with the old chaters)

```{r}
# Example code for coordinate transformations
```

## Raster Resampling and Reprojection
Adjusting the resolution of raster datasets or changing their projection...

```{r}
# Example code for raster resampling and reprojection
```
Merging and Appending Spatial Data
Combining multiple spatial datasets into one, either by stacking attributes (columns) or by appending features (rows).

```{r}
# Example code for merging and appending
```

## Handling Missing Data
Spatial datasets often have missing values. Here, we could discuss methods to identify, visualize, and impute missing data. (I would say, a important one)

```{r}
# Example code for handling missing data
```
## Spatial Data Cleaning
Detecting and correcting topological errors, such as slivers, gaps, or overlaps in polygon datasets.

```{r}
# Example code for spatial data cleaning
```

## Case Study: A Complete Spatial Data Wrangling Workflow
Maybe we could end it here with a Case Study to read that shows a workflow etc. Or just an Exercise?