# New Chapter: **Spatial Data Manipulation (see modified chapters)**

## Introduction
Here, we could briefly explain the difference between data wrangling and data manipulation in the spatial context. Highlight the importance of spatial data manipulation in GIS and spatial analysis workflows.

## Filtering Spatial Data
### Filtering sf Objects
Demonstrate how to filter spatial features in sf objects based on attribute conditions.

```{r}
# Example code for filtering sf objects
```
### Filtering Rasters
Show how to filter raster data based on cell values or spatial conditions.

```{r}
# Example code for filtering rasters
```

## Spatial Joins
Explain the concept of spatial joins and demonstrate how to join two spatial datasets based on their spatial relationship.

```{r}
# Example code for spatial joins
```

## Geometric Operations

###Intersect
Demonstrate how to find the common spatial elements between two datasets.

```{r}
# Example code for intersect operation
```

### Buffer
Show how to create a buffer zone around a spatial feature.

```{r}
# Example code for buffer operation
```

### Union and Difference
Explain and demonstrate the union and difference operations between two spatial datasets.

```{r}
# Example code for union and difference operations
```

## Coordinate Transformations and Reprojections
Discuss the importance of having datasets in the same coordinate system and demonstrate how to reproject spatial data.

```{r}
# Example code for coordinate transformations
```

## Raster Operations

### Resampling
Show how to change the resolution of raster datasets.

```{r}
# Example code for raster resampling
```

### Raster Algebra
Demonstrate basic raster algebra operations like addition, subtraction, and multiplication of raster layers.

```{r}
# Example code for raster algebra
```

## Comparative Analysis of Packages
Investigate and compare the performance (caching and stuff), ease of use, and capabilities of different packages like sf, terra, and sfdep for various spatial operations... (just an idea)

## Exercise ..