---
title: "Working with Spatial Raster Data using `terra`"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

The `terra` package in R provides tools for spatial data analysis. It's designed to be a successor to the `raster` package, offering faster performance and more consistent function names.

## Installation and Loading

Before we dive into the functionalities, let's ensure we have the `terra` package installed and loaded:

```{r setup}
library(terra)
```

## Creating a Raster
```{r}
# Create a raster with 5 rows, 5 columns, and initialize with random values
r <- rast(ncol=5, nrow=5)
values(r) <- runif(ncell(r))
plot(r)

```

## Reading and Writing Raster Data

```{r}
# Reading a raster file - insert a test TIF here
r <- rast("exampletif.tif")

# Writing a raster to a file and puts it right in your working directory if not assigned!
writeRaster(r, "output_file.tif")

```

## Raster Operations
terra provides a plethora of functions to perform operations on raster data.

### Basic Arithmetic
Performs some basic arithmetic operations on raster objects:
```{r}
r1 <- rast(ncol=5, nrow=5, vals=runif(25))
r2 <- rast(ncol=5, nrow=5, vals=runif(25))

result <- r1 + r2
plot(result)
```

### Check the CRS and Reproject it
```{r}
crs(r) 
r <- project(r, "+proj=longlat +datum=WGS84")

# if the raster does not hove a crs and you know the specific settings for it:
crs(r) <- "+proj=longlat +datum=WGS84"
r
```

### Aggregation
```{r}
# Aggregating the raster to half its resolution + error check
if (is.lonlat(r)) {
  # Store the original CRS
  original_crs <- crs(r)
  
  # Reproject to a suitable UTM zone (ajusted for Salzburg 33T)
  r <- project(r, "+proj=utm +zone=33T +datum=WGS84")
}

# Aggregating the raster to half its resolution
aggregated_r <- aggregate(r, fact=2)

# If we changed the CRS earlier, reproject back to the original CRS
if (exists("original_crs")) {
  aggregated_r <- project(aggregated_r, original_crs)
}

plot(aggregated_r)
```

## Conclusion

The `terra` package offers a comprehensive suite of tools for spatial raster data analysis in R. And much more is covered in Documentation see References.

## References

[r-spatial terra doc](https://rspatial.org/pkg/)
[cran - terra](https://cran.r-project.org/web/packages/terra/index.html)