---
title: "Global currency exchange report"
output:
  html_document: default
date: '2022-10-24'
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Global currencies to USD

The map shows currency exchange rates from national currencies to US Dollar. Data is retrieved by means of [Alpha Ventage API](https://www.alphavantage.co/){target="_blank"} in real-time at [foreign exchange opening hours](https://support.e-markets.nordea.com/hc/en-us/articles/360001333429-What-are-foreign-exchange-opening-hours-){target="_blank"}.

Alpha Ventage API provides free API service for up to 5 API requests per minute and 500 requests per day. To make sure that this limit is not exceeded, the request frequency is set to 1 request per 20 seconds. 

Currently, the code retrieves data for the first country in the list of alphabetically order countries, i.e. exchange rate of Afghan Afghani to USD. You may retrieve exchange rates of a larger number of countries (change variable `numcalls <- 1`).  

> Keep in mind that rendering time will increase with ~numcalls * 20sec.    

```{r currencymap, echo=FALSE, message=FALSE, warning=FALSE}
######This code retrieves today's currency exchange rates [USD to national currency] from Alpha Ventage and renders data in a world map#######
#Alpha Ventage API documentation: https://www.alphavantage.co/documentation/
#Documentation of library 'alphavantager' for R: https://www.rdocumentation.org/packages/alphavantager

library(maps)
library(alphavantager)
library(tidyverse)
library(leaflet)
library(tidyquant)
library(sfheaders)
library(htmltools)

#get currency codes from https://github.com/datasets/currency-codes
codes <- readr::read_csv("data/codes-all.csv")

rates <- c()
countries <- c()
currencies <- c()
numcalls <- 1      #set the number of national exchange rates to retrieve

# retrieve exchange rate for countries in range 1:numcalls as listed in table 'codes'
for (i in 1:length(codes$NumericCode[1:numcalls])){  
  
  #Make API call
  symbol <- paste(codes$AlphabeticCode[i], "/USD",sep="")
  base <- "https://www.alphavantage.co/query"
  apiKey <- alphavantager::av_api_key("5A80E9FDZ6KO093O")
  dat_alphaventage <- alphavantager::av_get(symbol, av_fun = "FX_DAILY")
  
  #Append API return to vectors
  if (length(dat_alphaventage) == 5){           #if: API call retrieved timestamp, open, high, low, close exchange rates
    
    today <- tail(dat_alphaventage$open, n=1)
    
    rates <- c(rates, today)
    countries <- c(countries, tolower(codes$Entity[i]))
    currencies <- c(currencies, codes$Currency[i])
      
    }
  else{                                         #else: API call returns only error message
    
    rates <- c(rates, NA)
    countries <- c(countries, tolower(codes$Entity[i]))
    currencies <- c(currencies, NA)
      
    }
    
    Sys.sleep(20)                               #sleep to make sure 5 calls/min is not exceeded
    
}

exchange_rates <- data.frame(countries, rates, currencies)      #combine country names and exchange rate vectors in a dataframe

#get world country dataframe and convert column to lower case letter for join
world_ctry <- ggplot2::map_data("world") %>%
  dplyr::mutate(region_lowercase = tolower(region))

#join API return and country dataframe
world_map <- world_ctry %>%
  dplyr::left_join(
    exchange_rates,
    by = c("region_lowercase" = "countries")
  )

#convert dataframe to SpatialPolygonsDataFrame for integration in Leaflet (see https://rstudio.github.io/leaflet/shapes.html)
world_map_sf <- sfheaders::sf_polygon(
  obj = world_map
  , x = "long"
  , y = "lat"
  , polygon_id = "group"
  , keep = TRUE
)

#create map in leaflet (see https://rstudio.github.io/leaflet/choropleths.html)
bins <- c(0.001, 0.01, 0.1, 0.2, 0.3, 0.4, 0.5, Inf)
pal <- colorBin("YlOrRd", domain = world_map_sf$rates, bins = bins)

labels <- sprintf("<strong>%s</strong><br/>1 %s is %g US Dollars </sup>",world_map_sf$region, world_map_sf$currencies, world_map_sf$rates) %>% 
  lapply(htmltools::HTML)

leaflet(data = world_map_sf) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~pal(rates),
              weight = 2,
              opacity = 1,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7,
              highlightOptions = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = TRUE),
              label = labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto")) %>%
  setView(lng = 0, lat = 10, zoom = 1.3) %>%
  leaflet::addLegend(pal = pal, values = ~rates, opacity = 0.7, title = NULL,
            position = "bottomright")
```

## Conclusions

A depreciation of a currency makes imports more expensive and leads to a demand shift away from foreign goods towards domestic goods [@towbin2013limits]. Conversely, strong currencies make imports more attractive.

## References
